Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O

### 2013-07-24 draft.tex
changed to elsarticle, added frontmatter, deleted big tables from the end

\documentclass{article}
\usepackage[top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage{amsmath}
\pagestyle{empty}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{lineno}
\usepackage{natbib}
\bibpunct{(}{)}{;}{a}{}{,}

\newenvironment{my_enumerate}{
\begin{enumerate}
  \setlength{\itemsep}{5pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}}{\end{enumerate}
}

@@

\subsection*{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}

\noindent Cummins, Renata C.$^1$; Finnegan, Seth$^2$; Fike, David A.$^3$; Eiler, John M.$^1$; Fischer, Woodward W.$^1$

\vspace{10pt}

\footnotesize

\noindent $^1$California Institute of Technology, Geological and Planetary Sciences, Pasadena, CA 91125

\noindent $^2$Department of Integrative Biology, University of California, 1005 Valley Life Sciences Bldg \#3140, Berkeley, CA 94720

\noindent $^3$Department of Earth and Planetary Sciences, Washington University, St. Louis, Missouri 63130

\normalsize

@@

\vspace{0.7cm} \noindent \large \textbf{Abstract} \normalsize

Much of what we know about the history of Earth's climate derives from the geologic record of carbonates. The oxygen isotopic composition (\deltao O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the \deltao O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the \deltao O of the water as well. The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set. These highly pristine samples ultimately provide a more reliable record of paleoclimate.

We tested the ability of carbonate clumped isotopes to constrain paleoclimate by measuring the clumped isotope composition of apparently well-preserved calcitic fossils from several Silurian-age (ca. 427 Ma) stratigraphic sections on the island of Gotland, Sweden. Variability in clumped isotope temperatures suggests differential preservation, complicating a resolved temperature history through these sections. Micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm ($30\pm 2$\degrees C) and similar in oxygen isotopic composition to a modern ice-free world (\deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time.

@@

\vspace{0.7cm} \noindent \large \textbf{Acknowledgements} \normalsize

@@

\bibliographystyle{garynat}
\bibliography{draft}

\begin{sidewaystable}[htbp]
\centering
\small
\begin{tabular}{ | l | c | c | c | c | c | c | c | c | c | c | }
\multirow{2}{*}{Sample} & Sample & Number of & \deltac C & \deltao O & $\Delta_{47}$ & Temperature & \deltao O water & [Fe] & [Mn] & [Sr] \\
 & location & replicates & ($\permil$ PDB) & ($\permil$ PDB) & ($\permil$) & (\degrees C) & ($\permil$ SMOW) & (ppm) & (ppm) & (ppm)\\ \hline
G1 -0.3 to -1.0 M1 & 1 & 3 & 2.575 & -5.277 & 0.589 & 38.7 $\pm$ 1.6 & -0.294 $\pm$ 0.30 & 4905 $\pm$ 29 & 327 $\pm$ 3 & 238 $\pm$ 2 \\
G1 -0.3 to -1.0 R1 & 1 & 3 & 2.319 & -5.320 & 0.596 & 36.8 $\pm$ 1.8 & -0.696 $\pm$ 0.33 & 337 $\pm$ 2 & 159 $\pm$ 1 & 332 $\pm$ 2 \\
G1 -1.2 to -0.8 B1 & 1 & 1 & 2.019 & -5.523 & 0.565 & 45.0 $\pm$ 2.7 & 0.602 $\pm$ 0.49 & 182 $\pm$ 5 & 66 $\pm$ 2 & 1351 $\pm$ 33 \\
G1 -3.2 to -3.0 R1 & 1 & 1 & 2.397 & -5.851 & 0.588 & 38.9 $\pm$ 3.1 & -0.833 $\pm$ 0.57 & 1330 $\pm$ 15 & 437 $\pm$ 3 & 428 $\pm$ 3 \\
G1 -3.4 to -3.2 B2 & 1 & 1 & 2.023 & -5.597 & 0.591 & 38.2 $\pm$ 1.5 & -0.713 $\pm$ 0.27 & 177 $\pm$ 3 & 53 $\pm$ 1 & 1507 $\pm$ 9 \\
G1 2.2 R1 & 1 & 1 & 3.734 & -5.190 & 0.587 & 39.1 $\pm$ 1.4 & -0.126 $\pm$ 0.25 & 172 $\pm$ 4 & 168 $\pm$ 1 & 431 $\pm$ 5 \\
G2 -0.2 to 0.0 R1 & 2 & 1 & 2.315 & -5.233 & 0.584 & 39.8 $\pm$ 2.5 & -0.046 $\pm$ 0.47 & 379 $\pm$ 3 & 153 $\pm$ 1 & 414 $\pm$ 5 \\
G2 -0.5 to -0.25 R1 & 2 & 1 & 2.208 & -5.351 & 0.592 & 37.9 $\pm$ 2.9 & -0.514 $\pm$ 0.53 & 887 $\pm$ 14 & 157 $\pm$ 1 & 600 $\pm$ 2 \\
G2 0.0 to 0.2 B1 & 2 & 1 & 2.399 & -5.692 & 0.584 & 40.0 $\pm$ 3.7 & -0.462 $\pm$ 0.68 & 302 $\pm$ 9 & 70 $\pm$ 1 & 1289 $\pm$ 23 \\
G2 0.5 to 0.6 B1 & 2 & 1 & 3.178 & -5.123 & 0.572 & 43.0 $\pm$ 2.8 & 0.639 $\pm$ 0.51 & 46 $\pm$ 3 & 64 $\pm$ 0.4 & 1447 $\pm$ 9 \\
G2 0.5 to 0.6 B2 & 2 & 1 & 2.925 & -5.347 & 0.602 & 35.4 $\pm$ 1.2 & -0.990 $\pm$ 0.23 & 38 $\pm$ 2 & 27 $\pm$ 1 & 1372 $\pm$ 3 \\
G3 0.0 B1 & 3 & 1 & 2.248 & -5.876 & 0.573 & 42.7 $\pm$ 2.1 & -0.166 $\pm$ 0.39 & 122 $\pm$ 1 & 48 $\pm$ 0.3 & 1127 $\pm$ 8 \\
G3 LVF B1 & 3 & 2 & 1.769 & -6.402 & 0.598 & 36.4 $\pm$ 1.9 & -1.850 $\pm$ 0.38 & 124 $\pm$ 2 & 65 $\pm$ 0.4 & 1493 $\pm$ 3 \\
G3 LVF B1 cement & 3 & 2 & 2.613 & -6.395 & 0.583 & 40.1 $\pm$ 2.9 & -1.155 $\pm$ 0.54 & 953 $\pm$ 9 & 663 $\pm$ 8 & 259 $\pm$ 4 \\
G3 LVF B1 micrite & 3 & 3 & 1.768 & -5.998 & 0.581 & 40.7 $\pm$ 1.5 & -0.647 $\pm$ 0.29 & 5091 $\pm$ 14 & 457 $\pm$ 2 & 226 $\pm$ 3 \\
G3 LVF B5 & 3 & 1 & 1.792 & -5.453 & 0.610 & 33.4 $\pm$ 1.6 & -1.479 $\pm$ 0.30 &  $\pm$  &  $\pm$  &  $\pm$  \\
G3 LVF B5 micrite & 3 & 2 & 1.628 & -5.723 & 0.586 & 39.4 $\pm$ 2.7 & -0.614 $\pm$ 0.49 & 4153 $\pm$ 22 & 449 $\pm$ 3 & 170 $\pm$ 2 \\
G3 LVF R1 & 3 & 2 & 2.075 & -5.440 & 0.577 & 41.8 $\pm$ 2.7 & 0.112 $\pm$ 0.48 & 724 $\pm$ 7 & 161 $\pm$ 2 & 435 $\pm$ 4 \\
G3 LVF R2 & 3 & 2 & 5.942 & -5.368 & 0.595 & 37.2 $\pm$ 2.0 & -0.658 $\pm$ 0.37 & 69 $\pm$ 3 & 211 $\pm$ 1 & 309 $\pm$ 5 \\
G4 2.0 to 3.0 B1 & 4 & 1 & 4.223 & -6.092 & 0.580 & 40.9 $\pm$ 2.8 & -0.712 $\pm$ 0.51 &  nd  &  nd  &  nd  \\
G4 2.0 to 3.0 bryozoan1 & 4 & 1 & 4.582 & -5.604 & 0.589 & 38.6 $\pm$ 3.8 & -0.635 $\pm$ 0.71 &  nd  &  nd  &  nd  \\
G4 2.0 to 3.0 crinoid1 & 4 & 1 & 5.033 & -6.216 & 0.564 & 45.2 $\pm$ 3.3 & -0.050 $\pm$ 0.58 &  nd  &  nd  &  nd  \\
G5 1.6 to 2.2 gastropod1 & 5 & 1 & 5.589 & -6.210 & 0.598 & 36.4 $\pm$ 2.4 & -1.667 $\pm$ 0.45 &  nd  &  nd  &  nd  \\
G5 1.6 to 2.2 ostracod1 & 5 & 1 & 4.425 & -4.839 & 0.585 & 39.6 $\pm$ 3.3 & 0.309 $\pm$ 0.60 &  nd  &  nd  &  nd  \\
G6 B234 & 6 & 1 & 4.692 & -5.152 & 0.591 & 38.1 $\pm$ 1.4 & -0.286 $\pm$ 0.26 &  nd  &  nd  &  nd  \\
G6 crinoid1 & 6 & 1 & 5.457 & -6.380 & 0.535 & 53.3 $\pm$ 4.0 & 1.187 $\pm$ 0.68 &  nd  &  nd  &  nd  \\
G7 9.0 R1 & 7 & 2 & 5.336 & -5.495 & 0.575 & 42.2 $\pm$ 2.5 & 0.135 $\pm$ 0.45 & 121 $\pm$ 3 & 192 $\pm$ 1 & 86 $\pm$ 6 \\
G7 9.0 R2 & 7 & 1 & 5.173 & -5.698 & 0.551 & 48.9 $\pm$ 2.6 & 1.120 $\pm$ 0.45 &  nd  &  nd  &  nd  \\
G8 0.0 to 1.2 R1 & 8 & 1 & -0.507 & -5.208 & 0.593 & 37.7 $\pm$ 2.7 & -0.419 $\pm$ 0.50 & 187 $\pm$ 4 & 78 $\pm$ 1 & 339 $\pm$ 3 \\
G8 1.2 to 2.0 M1 & 8 & 1 & -0.768 & -7.093 & 0.561 & 46.0 $\pm$ 2.2 & -0.801 $\pm$ 0.39 & 3635 $\pm$ 13 & 248 $\pm$ 1 & 24 $\pm$ 6 \\
G8 1.2 to 2.0 R1 & 8 & 2 & -0.082 & -5.022 & 0.553 & 48.2 $\pm$ 2.7 & 1.676 $\pm$ 0.47 & 128 $\pm$ 4 & 78 $\pm$ 1 & 191 $\pm$ 3 \\
\end{tabular} 
\caption{Isotopic composition, clumped isotope temperature, calculated \deltao O of water and bulk trace metal concentrations of all the samples analyzed in this study. Errors are standard error of the mean for all acqusitions of that sample. $\text{nd}=\text{no data.}$}
\label{results}
\end{sidewaystable}

\begin{sidewaystable}[htbp]
\centering
\small
\begin{tabular}{ | l | c | c | c | c | c | c | c | c | c | c | }
\multirow{2}{*}{Sample} & Sample & Number of & \deltac C & \deltao O & $\Delta_{47}$ & Temperature & \deltao O water & [Fe] & [Mn] & [Sr] \\
 & location & replicates & ($\permil$ PDB) & ($\permil$ PDB) & ($\permil$) & (\degrees C) & ($\permil$ SMOW) & (ppm) & (ppm) & (ppm)\\ \hline
G11 0.0 to 0.2 R1 & 11 & 1 & 3.164 & -4.847 & 0.599 & 36.1 $\pm$ 2.4 & -0.352 $\pm$ 0.46 & 116 $\pm$ 4 & 140 $\pm$ 1 & 769 $\pm$ 4 \\
G11 0.2 to 0.4 B2 & 11 & 1 & 2.836 & -5.360 & 0.587 & 39.2 $\pm$ 3.6 & -0.289 $\pm$ 0.67 & 252 $\pm$ 6 & 59 $\pm$ 1 & 1424 $\pm$ 15 \\
G11 0.2 to 0.4 B6 & 11 & 1 & 2.454 & -6.146 & 0.605 & 34.6 $\pm$ 2.3 & -1.939 $\pm$ 0.43 & 89 $\pm$ 2 & 107 $\pm$ 1 & 1450 $\pm$ 8 \\
G11 0.2 to 0.4 R1 & 11 & 1 & 3.169 & -5.495 & 0.569 & 43.9 $\pm$ 2.2 & 0.440 $\pm$ 0.40 & 487 $\pm$ 7 & 346 $\pm$ 2 & 481 $\pm$ 4 \\
G11 0.4 to 0.6 B1 & 11 & 1 & 2.766 & -5.439 & 0.596 & 36.9 $\pm$ 1.6 & -0.788 $\pm$ 0.30 & 58 $\pm$ 1 & 40 $\pm$ 0.2 & 1476 $\pm$ 4 \\
G11 0.6 B1 & 11 & 1 & 2.927 & -5.269 & 0.588 & 38.9 $\pm$ 1.8 & -0.242 $\pm$ 0.33 & 59 $\pm$ 2 & 59 $\pm$ 1 & 1627 $\pm$ 11 \\
G11 0.6 to 0.8 M1 & 11 & 1 & 2.431 & -6.468 & 0.610 & 33.5 $\pm$ 2.0 & -2.472 $\pm$ 0.39 & 20948 $\pm$ 90 & 288 $\pm$ 7 & 79 $\pm$ 3 \\
G11 0.6 to 0.8 R1 & 11 & 1 & 3.447 & -4.886 & 0.596 & 36.8 $\pm$ 3.5 & -0.258 $\pm$ 0.65 & 75 $\pm$ 4 & 68 $\pm$ 1 & 503 $\pm$ 6 \\
G12 B3 & 12 & 1 & 2.803 & -5.757 & 0.558 & 46.9 $\pm$ 1.8 & 0.707 $\pm$ 0.32 &  nd  &  nd  &  nd  \\
G12 B4 & 12 & 1 & 4.397 & -5.179 & 0.581 & 40.8 $\pm$ 3.7 & 0.191 $\pm$ 0.67 & 64 $\pm$ 3 & 41 $\pm$ 1 & 1358 $\pm$ 4 \\
G12 R1 & 12 & 3 & 2.381 & -5.221 & 0.604 & 35.0 $\pm$ 1.6 & -0.935 $\pm$ 0.31 & 601 $\pm$ 6 & 124 $\pm$ 1 & 412 $\pm$ 4 \\
G12 R1 micrite & 12 & 1 & 2.102 & -5.048 & 0.568 & 44.2 $\pm$ 3.8 & 0.939 $\pm$ 0.69 &  nd  &  nd  &  nd  \\
G12 R2 & 12 & 1 & 2.043 & -5.611 & 0.582 & 40.6 $\pm$ 2.8 & -0.287 $\pm$ 0.52 & 1474 $\pm$ 18 & 417 $\pm$ 2 & 342 $\pm$ 2 \\
G13 1.0 to 2.0 R1 & 13 & 1 & -0.568 & -4.974 & 0.582 & 40.4 $\pm$ 4.4 & 0.320 $\pm$ 0.81 & 26 $\pm$ 4 & 107 $\pm$ 1 & 200 $\pm$ 3 \\
G13 10.5 to 10.8 B1 & 13 & 1 & -0.598 & -4.986 & 0.626 & 29.6 $\pm$ 2.4 & -1.741 $\pm$ 0.48 & 32 $\pm$ 5 & 16 $\pm$ 2 & 1372 $\pm$ 109 \\
G13 10.5 to 10.8 B1 leftovers & 13 & 1 & -0.722 & -5.207 & 0.579 & 41.3 $\pm$ 4.4 & 0.251 $\pm$ 0.81 &  nd  &  nd  &  nd  \\
G13 10.5 to 10.8 M1 & 13 & 1 & -1.496 & -6.298 & 0.547 & 50.0 $\pm$ 2.7 & 0.703 $\pm$ 0.46 & 3487 $\pm$ 7 & 277 $\pm$ 1 & 31 $\pm$ 4 \\
G13 10.5 to 10.8 R1 tip & 13 & 2 & -0.564 & -4.810 & 0.594 & 37.2 $\pm$ 2.0 & -0.097 $\pm$ 0.37 &  nd  &  nd  &  nd  \\
G13 10.5 to 10.8 R1 top & 13 & 2 & -1.044 & -5.829 & 0.583 & 40.3 $\pm$ 1.5 & -0.557 $\pm$ 0.28 & 571 $\pm$ 65 & 193 $\pm$ 13 & 232 $\pm$ 15 \\
G13 31.4 to 32.4 B1 & 13 & 3 & -0.316 & -5.082 & 0.592 & 37.9 $\pm$ 1.8 & -0.243 $\pm$ 0.34 & 141 $\pm$ 3 & 27 $\pm$ 0.3 & 1436 $\pm$ 11 \\
G13 31.4 to 32.4 M1 & 13 & 1 & -0.778 & -5.965 & 0.570 & 43.5 $\pm$ 2.6 & -0.109 $\pm$ 0.47 & 3671 $\pm$ 10 & 237 $\pm$ 1 & 42 $\pm$ 3 \\
G13 31.4 to 32.4 R1 & 13 & 1 & -0.809 & -5.484 & 0.535 & 53.4 $\pm$ 4.0 & 2.103 $\pm$ 0.68 & 96 $\pm$ 7 & 143 $\pm$ 1 & 118 $\pm$ 6 \\
G13 33.5 B1 bad & 13 & 1 & -0.556 & -5.253 & 0.576 & 42.1 $\pm$ 3.3 & 0.361 $\pm$ 0.61 & 107 $\pm$ 1 & 46 $\pm$ 0.3 & 1320 $\pm$ 5 \\
G13 33.5 B1 leftovers & 13 & 1 & -0.191 & -5.338 & 0.575 & 42.4 $\pm$ 3.0 & 0.321 $\pm$ 0.54 & 19 $\pm$ 5 & 14 $\pm$ 2 & 983 $\pm$ 109 \\
G13 33.5 B1 perfect & 13 & 1 & -0.159 & -5.407 & 0.596 & 36.9 $\pm$ 3.0 & -0.756 $\pm$ 0.56 &  nd  &  nd  &  nd  \\
G13 35.6 to 36.0 B1 & 13 & 1 & -0.288 & -5.397 & 0.580 & 41.0 $\pm$ 1.9 & 0.011 $\pm$ 0.35 & 39 $\pm$ 1 & 14 $\pm$ 0.1 & 1657 $\pm$ 9 \\
G13 5.0 to 6.0 B3 & 13 & 1 & -0.216 & -5.614 & 0.576 & 42.1 $\pm$ 3.5 & -0.001 $\pm$ 0.64 &  nd  &  nd  &  nd  \\
G13 5.0 to 6.0 B3 cement & 13 & 1 & -0.909 & -7.169 & 0.564 & 45.3 $\pm$ 2.2 & -0.996 $\pm$ 0.39 &  nd  &  nd  &  nd  \\
G13 6.5 to 7.5 M1 & 13 & 1 & -0.118 & -5.206 & 0.559 & 46.7 $\pm$ 3.9 & 1.229 $\pm$ 0.69 & 5708 $\pm$ 34 & 289 $\pm$ 3 & 0 $\pm$ 2 \\
G13 6.5 to 7.5 R1 & 13 & 1 & 0.180 & -4.728 & 0.536 & 53.3 $\pm$ 2.4 & 2.847 $\pm$ 0.41 & 32 $\pm$ 2 & 80 $\pm$ 1 & 204 $\pm$ 6 \\
G13 UF B1 & 13 & 3 & -0.347 & -5.416 & 0.598 & 36.4 $\pm$ 1.4 & -0.857 $\pm$ 0.28 & 27 $\pm$ 1 & 35 $\pm$ 0.5 & 1564 $\pm$ 7 \\
G13 UF B1 cement & 13 & 2 & -0.868 & -7.730 & 0.578 & 41.4 $\pm$ 2.0 & -2.259 $\pm$ 0.37 & 234 $\pm$ 2 & 266 $\pm$ 2 & 100 $\pm$ 1 \\
G13 UF B1 micrite & 13 & 2 & -1.024 & -6.330 & 0.543 & 51.1 $\pm$ 2.2 & 0.852 $\pm$ 0.38 & 1428 $\pm$ 6 & 251 $\pm$ 2 & 126 $\pm$ 2 \\
\end{tabular} 
\end{sidewaystable}

>>>>

\documentclass[preprint, authoryear]{elsarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{enumerate}
\usepackage{wasysym}
\usepackage{natbib}
\usepackage{lineno}

@@

\begin{frontmatter}

\title{Carbonate clumped isotope constraints on Silurian ocean temperature and \deltao O}

\author[caltech]{Renata C. Cummins\corref{cor1}}
\ead{renatacummins@gmail.com}

\author[berkley]{Seth Finnegan}

\author[washU]{David A. Fike}

\author[caltech]{John M. Eiler}

\author[caltech]{Woodward W. Fischer}

\cortext[cor1]{Corresponding author}

\address[caltech]{California Institute of Technology, Geological and Planetary Sciences, MC 100-23, Pasadena, CA 91125}

\address[berkley]{Department of Integrative Biology, University of California, 1005 Valley Life Sciences Bldg \#3140, Berkeley, CA 94720}

\address[washU]{Department of Earth and Planetary Sciences, Washington University, St. Louis, Missouri 63130}

\begin{abstract}
Much of what we know about the history of Earth's climate derives from the geologic record of carbonates. The oxygen isotopic composition (\deltao O) of carbonates has been widely used as a proxy for past ocean temperatures, but application of this proxy to deep geologic time is complicated by diagenetic alteration and uncertainties in the \deltao O of seawater in the past. Carbonate clumped isotope thermometry provides an independent estimate of the temperature of the water from which a carbonate precipitated, allowing direct calculation of the \deltao O of the water as well. The clumped isotope composition of carbonates is also highly sensitive to diagenetic alteration, enabling identification of the most pristine samples from within a sample set. These highly pristine samples ultimately provide a more reliable record of paleoclimate.

We tested the ability of carbonate clumped isotopes to constrain paleoclimate by measuring the clumped isotope composition of apparently well-preserved calcitic fossils from several Silurian-age (ca. 427 Ma) stratigraphic sections on the island of Gotland, Sweden. Variability in clumped isotope temperatures suggests differential preservation, complicating a resolved temperature history through these sections. Micro-scale studies of calcite fabric and trace metal composition using electron backscatter diffraction and electron microprobe analysis point to a subset of relatively pristine samples, which indicate that Silurian tropical oceans were likely warm ($30\pm 2$\degrees C) and similar in oxygen isotopic composition to a modern ice-free world (\deltao O$_{\text{SMOW}}$ of $-1.7\pm 0.5\permil$). This result joins the growing body of evidence that suggests the $\delta^{18}$O of Earth's surface waters has remained broadly constant through time. 
\end{abstract}

\end{frontmatter}

@@

@@

\section*{Acknowledgements}

@@

\section*{References}
\bibliographystyle{my-harv}
\bibliography{draft}
